### 2012-02-13 doc/theses/mikula/analysis_of_xquery.tex
Master thesis: In progress.

\chapter{Analysis of XQuery} \label{CHAPTER_analysis_of_xquery}
\todo[inline]{Někde by se mělo říct, kterou verzi uvažujeme, jaké obecně jsou, co je třeba XQuery Core apod.}
This chapter discusses selected constructs of XQuery language and denotes how they could be exploited in the XML schema inference process. It is divided into sections by particular domains of the inference.

Most of sample queries in this chapter are taken from \cite{Walmsley:2007:XQU:1205865} and \cite{Marchiori:07:XQU}, with or without modifications, as these were the main sources for this XQuery analysis.
\todo[inline]{Overit si, ze mozem pouzivat tieto prevzate priklady.}

\section{Structure of XML documents}
Most of queries can be exploited to obtain some information about a structure of respective XML documents. The structure of XML documents captures elements from these documents along with their names, attributes and their organization. What the root elements in these documents are, which elements can be contained within a certain element, whether they are optional or mandatory, etc.

Path expressions without predicates which use only child axis are the simpliest example of such queries. Path expression \texttt{/bib/book/author} indicates that element \texttt{bib} is the outermost element and it contains one or more elements \texttt{book} and these contain one or more elements \texttt{author}.

Additional path expressions \texttt{/bib/book/title}, \texttt{//author/name} indicate that element \texttt{book} also contains element \texttt{title} and element \texttt{author} contains element \texttt{name}. The latter one uses also descendant axis. That means the query considers all elements \texttt{author} in the document, thus it is a hint that maybe there are several elements named \texttt{author} but with different absolute paths and all of them contain an element \texttt{name}.

Besides elements, attributes can be processed exactly in the same way. Path expression \texttt{/bib/book/@ISBN} indicates that element \texttt{book} has attribute \texttt{ISBN}.

However, statements of this kind are necessary for the XML schema inference, their obtaining from queries is not significant due to the following reason. These statements could be determined directly from the XML documents and it could be easily done in a more convenient way. Queries may not cover the whole relevant content of the documents. Let \texttt{E} be an element occurring in an XML document and let \texttt{Q} be an XQuery query. Consider the following cases:

\begin{enumerate}[1)]
\item \texttt{E} is directly mentioned in \texttt{Q}. For example elements \texttt{bib}, \texttt{book} and \texttt{author} are mentioned in path expression \texttt{/bib/book/author}.
\item \texttt{E} is not mentioned in \texttt{Q} but it does occur in a result of the evaluation of \texttt{Q}. For example path expression \texttt{/bib/book/author} returns elements \texttt{author} along with their subelements \texttt{name}, \texttt{birthdate} and \texttt{nationality}.
\item \texttt{E} is not mentioned in \texttt{Q} nor it occurs in the result but it is processed by the evaluation of \texttt{Q}. For example the evaluation of path expression \texttt{//author} processes amongst others elements \texttt{bib} and \texttt{book}, in a search for elements \texttt{author}.
\item \texttt{E} occurs in a part of the XML document not related to a query at all.
\end{enumerate}

@@

\section{Occurrence Count of Elements}
Some XQuery constructs indicate multiplicity of a particular element or limit the element to occur at most once. Consider the following query assuming variable \texttt{\$book1} is bound to a certain \texttt{book} element.

@@

Function \texttt{count()} returns number of items in a provided sequence. If the sequence is a sequence of elements, the count of these elements will probably not be limited to one. The sample query indicates that the root element \texttt{book} can contain more than one element \texttt{section}.
An exception is usage of function \texttt{count()} in a predicate in expressions where it is used to determine if count of some nodes is greater than zero. Often, this form is used to test presence of a certain node instead of actual counting of its occurrences. In this case, the node could be still limited to occur at most once.

@@

In this query, number of elements \texttt{color} in element \texttt{product} with attribute \texttt{id} equal to 5 has to be at least one, otherwise an error is raised upon execution. If we assume that this query is written correctly, with common sense and it should not raise the error, we can infer that element \texttt{product} has to contain one or more elements \texttt{color}.

@@

Alike the previous example, an expression in \texttt{order by} clause can be evaluated to at most one item or the \emph{type error} is raised. Therefore, every element \texttt{item} contains zero or one element \texttt{num} but no more.

@@

\subsection{XML Schema Built-in Types}
XML Schema built-in types of elements and attributes can be inferred from the XML documents by analysing their content. Since the count of built-in types is 44 and inheritance is involved, such analysis may be imprecise, especially when a large enough set of XML documents is not available. This is when a set of XQuery queries may come in handy. For example, consider the following occurrences of element \texttt{a}.

@@

These three occurrences are not sufficient enough to determine the accurate type of element \texttt{a}, because values \texttt{1}, \texttt{6}, \texttt{18} are valid values of several types: \texttt{decimal}, \texttt{integer}, \texttt{byte}, \texttt{short}, \texttt{unsignedInt}, \texttt{positiveInteger} and many others. The following text introduces some possibilities of a type inference from queries.

If a value of an element or attribute is used in an expression, then this expression could be often exploited to determine a type of the value. Comparing the value to another value of a known type and supplying the value to a function call as an argument of particular type are examples of these expressions as shown below.

@@

\subsection{Enumeration}
In several cases, XQuery \emph{if-then-else} construct may be used to branch the execution of query by all possible values of a certain variable. This is equivalent of \emph{switch} construct from other programming languages. When the control variable is bound to some element or attribute, type of this node can be inferred as enumeration and its individual values can be determined as well.

@@

The following examples are queries that could be processed by a similar method but the actual method will not use them.

@@

A common three-way join.

@@

\subsection{Negative Statements about Uniqueness} \label{Negative_uniqueness}
In many cases a statement refusing uniqueness of element or attribute values could be inferred. Such statements may be helpful in combination with other methods in its process of making a decision whether a particular element or attribute is unique or not.

@@

Usage of \texttt{for} construct indicates that a sequence of \texttt{book} elements which satisfy the condition in \emph{where} clause is expected. It is a condition composed of two single conditions joined by \emph{and} logical operator. Therefore, in order to satisfy the whole condition, both single conditions must be satisfied as well. Thus, it is expected that several elements satisfy each of the single conditions.

The first of them is test of equality of book's subelement \texttt{publisher} to a string literal. Based on the expectation, element \texttt{publisher} cannot be unique. However, the second condition is greater-than comparison of \texttt{year} attribute to an integer literal, it cannot be inferred whether this attribute is unique or not. The reason is that even if it was unique, there still might be more than one \texttt{book} element meeting this condition. Also, any statement, positive nor negative, cannot be inferred about \texttt{title} subelement.

@@

@@

For each \texttt{product} element, new element \texttt{prod} with attribute \texttt{xml:id} is created. Since attribute \texttt{xml:id} is supposed to be unique and there is direct transformation of the values of \texttt{number} elements to the values of attribute \texttt{xml:id}, it is very likely that element \texttt{number} is unique in the source data.

@@

\begin{verbatim}
let $i := //incision)[2]
for $a in //action)[. >> $i]
return $a//instrument
\end{verbatim}

@@

\subsection{Intermediate XML structure}
Intermediate XML structure are XML data that are not read from input XML documents nor created as an output of queries but they are somehow used by the queries. Objective of this section is intermediate XML structure created directly in the queries. It may serve to various purposes and it can be created in various ways, hard-coded in queries, computed from the input XML documents to simplify their structure. The former is demonstrated in the following example.

>>>>

svn diff\chapter{Analysis of XQuery} \label{CHAPTER_analysis_of_xquery}
\todo[inline]{Někde by se mělo říct, kterou verzi uvažujeme, jaké obecně jsou, co je třeba XQuery Core apod.}
This chapter discusses selected constructs of XQuery language and denotes how they could be exploited in the XML schema inference process. It is divided into sections by particular domains of the inference.

Most of sample queries in this chapter are taken from \cite{Walmsley:2007:XQU:1205865} and \cite{Marchiori:07:XQU}, with or without modifications, as these were the main sources for this XQuery analysis.

\section{Structure of XML documents}
Most of queries can be exploited to obtain some information about the structure of respective XML documents. The structure of XML documents captures elements from these documents along with their names, attributes and their organization. What the root elements in these documents are, which elements can be contained within a certain element, whether they are optional or mandatory, etc.

Path expressions without predicates which use only child axis are the simpliest example of such queries. For example, path expression \texttt{/bib/book/author} indicates that element \texttt{bib} is the root element and it contains one or more elements \texttt{book} and these contain one or more elements \texttt{author}.

Additional path expressions \texttt{/bib/book/title}, \texttt{//author/name} indicate that element \texttt{book} also contains element \texttt{title} and element \texttt{author} contains element \texttt{name}. The latter one uses also descendant axis. That means the query considers all elements \texttt{author} in the document, thus it is a hint that maybe there are several elements named \texttt{author} but with different absolute paths and some of them contain an element \texttt{name}.

Besides elements, attributes can be processed exactly in the same way. Path expression \texttt{/bib/book/@ISBN} indicates that element \texttt{book} can have attribute \texttt{ISBN}.

Statements of this kind are necessary for the XML schema inference, however their obtaining from queries is not significant due to the following reason. These statements could be determined directly from the XML documents and it could be easily done in a more convenient way. In addition, queries may not cover the whole relevant content of the documents. Let \texttt{E} be an element occurring in an XML document and let \texttt{Q} be an XQuery query. Consider the following cases:

\begin{enumerate}[1)]
\item \texttt{E} is directly mentioned in \texttt{Q}. For example elements \texttt{bib}, \texttt{book} and \texttt{author} are mentioned in path expression \texttt{/bib/book/author}.
\item \texttt{E} is not mentioned in \texttt{Q} but it does occur in a result of the evaluation of \texttt{Q}. For example path expression \texttt{/bib/book/author} returns elements \texttt{author} along with their subelements \texttt{name}, \texttt{birthdate} and \texttt{nationality}.
\item \texttt{E} is neither mentioned in \texttt{Q}, nor it occurs in the result, but it is processed by the evaluation of \texttt{Q}. For example the evaluation of path expression \texttt{//author} processes amongst others elements \texttt{bib} and \texttt{book}, in a search for elements \texttt{author}.
\item \texttt{E} occurs in a part of the XML document not related to \texttt{Q} at all.
\end{enumerate}

@@

\section{Number of Occurrences of Elements}
Some XQuery constructs indicate multiplicity of a particular element or limit the element to occur at most once. Consider the following query assuming variable \texttt{\$book1} is bound to a certain \texttt{book} element.

@@

Function \texttt{count()} returns the number of items in a provided sequence. If the sequence is a sequence of elements, the number of these elements will probably not be limited to one. The sample query indicates that the root element \texttt{book} can contain more than one element \texttt{section}.
An exception is usage of function \texttt{count()} in a predicate in expressions where it is used to determine if count of some nodes is greater than zero. Often, this form is used to test presence of a certain node instead of actual counting of its occurrences. In this case, the node could be still limited to occur at most once.

@@

In this query, the number of elements \texttt{color} in element \texttt{product} with attribute \texttt{id} equal to 5 has to be at least one, otherwise an error is raised upon execution. If we assume that this query is written correctly, with common sense and it should not raise the error, we can infer that element \texttt{product} has to contain one or more elements \texttt{color}.

@@

Alike the previous example, an expression in \texttt{order by} clause can be evaluated to at most one item or the \emph{type error} is raised. Therefore, every element \texttt{item} contains zero or one element \texttt{num} but not more.

@@

\subsection{XML Schema Built-in Types}
XML Schema built-in types of elements and attributes can be inferred from the XML documents by analysing their content. Since the number of built-in types is 44 and inheritance is involved, such analysis may be imprecise, especially when a large enough set of XML documents is not available. This is a case when a set of XQuery queries may bring optimizations. For example, consider the following occurrences of element \texttt{a}.

@@

These three occurrences are not sufficient enough to determine the accurate type of element \texttt{a}, because values \texttt{1}, \texttt{6}, \texttt{18} are valid values of several types: \texttt{decimal}, \texttt{integer}, \texttt{byte}, \texttt{short}, \texttt{unsignedInt}, \texttt{positiveInteger} and many others.

If a value of an element or attribute is used in an expression, then this expression could be often exploited to determine a type of the value. Comparing the value to another value of a known type and supplying the value to a function call as an argument of a particular type are examples of these expressions as shown below.

@@

\subsection{Enumeration}
In several cases, XQuery \emph{if-then-else} construct may be used to branch the execution of query by all possible values of a certain variable. This is equivalent to \emph{switch} construct from other programming languages. When the control variable is bound to some element or attribute, type of this node can be inferred as enumeration and its individual values can be determined as well.

@@

The following examples are queries that could be processed by a similar method, but the actual method will not use them.

@@

A common three-way join. \todo{Tady asi něco chybí.}

@@

\subsection{Negative Statements about Uniqueness} \label{Negative_uniqueness}
In many cases a statement refusing uniqueness of element or attribute values can be inferred. Such statements may be helpful in combination with other methods in its process of making a decision whether a particular element or attribute is unique or not.

@@

Usage of \texttt{for} construct indicates that a sequence of \texttt{book} elements which satisfy the condition in \emph{where} \todo{tady se nám nějak změnil styl} clause is expected. It is a condition composed of two single conditions joined by \emph{and} logical operator. Therefore, in order to satisfy the whole condition, both single conditions must be satisfied as well. Thus, it is expected that several elements satisfy each of the single conditions.

The first of them is a test of equality of book's subelement \texttt{publisher} to a string literal. Based on the expectation, element \texttt{publisher} cannot be unique. However, the second condition is greater-than comparison of \texttt{year} attribute to an integer literal, it cannot be inferred whether this attribute is unique or not. The reason is that even if it was unique, there still might be more than one \texttt{book} element meeting this condition. Also, any statement, positive nor negative, cannot be inferred about \texttt{title} subelement.

@@

\todo[inline]{Tady to chce ějaký komentář.}

@@

For each \texttt{product} element, new element \texttt{prod} with attribute \texttt{xml:id} is created. Since attribute \texttt{xml:id} is supposed to be unique and there is a direct transformation of the values of \texttt{number} elements to the values of attribute \texttt{xml:id}, it is very likely that element \texttt{number} is unique in the source data.

@@

\begin{verbatim}
let $i := //incision[2]
for $a in //action[. >> $i]
return $a//instrument
\end{verbatim}

@@

\subsection{Intermediate XML structure}
Intermediate XML structure represents XML data that are neither read from input XML documents nor created as an output of queries but they are somehow used by the queries. Objective of this section is intermediate XML structure created directly in the queries. It may serve to various purposes and it can be created in various ways, hard-coded in queries, computed from the input XML documents to simplify their structure. The former is demonstrated in the following example.
